\Large{IMPLEMENTASI FILTER SPASIAL LINEAR PADA VIDEO \textit{STREAM} MENGGUNAKAN \textit{FPGA HARDWARE ACCELERATOR}}

### 2021-05-18 include/halaman-abstract.tex
abstract

Various kinds of accelerators have been developed to improve performance and energy efficiency to handle heavy computations, one of which is FPGA. FPGA is capable of handling such a heavy computational load that it can be used for Digital Signal Processing, Image Processing, Neural Networks, etc. In this study, the authors tried to examine the performance of the ARM processor and the FPGA on the Xilin PYNQ Z2 FPGA Development Board in applying a linear spatial filter to the video stream. Kernel filters used in this study are the average blur, Gaussian blur, Laplacian, sharpen, Sobel horizontal, and Sobel vertical. The parameters used to measure the performance of ARM processors and FPGAs are runtime, frame rate (FPS), CPU usage, memory usage, resident memory (RES), shared memory (SHR), and virtual memory (VIRT). The average computation time required to apply linear spatial filters of 200 frames with an ARM processor is 29.06 seconds, while the average FPGA takes only 3.32 seconds. The filtered video with the ARM processor gets an average of 6.95 fps while the FPGA average is 60.37 fps. CPU usage on FPGAs is slightly lower than ARM processors. In general, the use of memory, resident memory, shared memory, and virtual memory on ARM processors and FPGAs is not much different.

>>>>

Various kinds of accelerators have been developed to improve performance and energy efficiency to handle heavy computations, one of which is FPGA. FPGA is capable of handling such a heavy computational load that it can be used for Digital Signal Processing, Image Processing, Neural Networks, etc. In this study, the authors tried to examine the performance of the ARM processor and the FPGA on the Xilin PYNQ Z2 FPGA Development Board in applying a linear spatial filter to the video stream. Kernel filters used in this study are the average blur, Gaussian blur, Laplacian, sharpen, Sobel horizontal, and Sobel vertical. The parameters used to measure the performance of ARM processors and FPGAs are runtime, frame rate (FPS), CPU usage, memory usage, resident memory (RES), shared memory (SHR), and virtual memory (VIRT). The average computation time required to apply linear spatial filters to 200 frames with an ARM processor is 29.06 seconds, while the average FPGA takes only 3.32 seconds. Compute time with FPGA is 88.85\% better than ARM processor. The filtered video with the ARM processor gets an average of 6.95 fps while the FPGA average is 60.37 fps. FPS with FPGA is 88.49\% better than ARM processor. CPU usage on FPGA is 14.89\% better, memory usage on FPGA is 2.02\% better, usage of resident memory is 2.07\% better, and usage of shared memory is 4.08\% better than ARM processor. While the use of virtual memory on ARM processors is 0.03\% better than FPGA.

### 2021-05-18 include/kata-pengantar.tex
kata pengantar
abstract

Ucapat terima kasih dan apresiasi yang tak terhingga kepada kedua orang tua penulis bapak \textbf{Sudarmin} dan ibu \textbf{Yuli Hadiyanti} yang tak kenal lelah dalam memanjatkan doa serta memberikan nasihat dan motivasi kepada penulis. Tak lupa juga kepada saudara-saudara penulis \textbf{Fitri Handayani}, \textbf{Tri Novianti}, \textbf{Jumadil Yusuf}, \textbf{Muhammad Fitrah}, \textbf{Adam Ramadhan} yang selalu menjadi motivasi bagi penulis untuk terus melangkah maju.

@@

\begin{enumerate}[topsep=0pt,itemsep=0pt,partopsep=0pt, parsep=0pt]
    \item Rektor Universitas Hasanuddin, Ibu \textbf{Prof. Dr. Dwia Aries Tina Pulubuhu} beserta jajarannya.
    \item Dekan Fakultas Matematika dan Ilmu Pengetahuan Alam, \textbf{Dr. Eng. Amiruddin} beserta jajarannya.
    \item Ketua Departemen Matematika FMIPA, \textbf{Dr. Nurdin, S.Si., M.Si}, dan juga \textbf{Dr. Muhammad Hasbi, M.Sc} sebagai ketua Program Studi Sistem Informasi Universitas Hasanuddin.
    \item Bapak \textbf{Dr. Eng. Armin Lawi, S.Si., M.Eng.} sebagai pembimbing utama yang telah banyak memberikan arahan, ide, motivavsi serta dukungan kepada penulis dalam banyak hal.
    \item Almarhum bapak \textbf{Dr. Diaraya, M.Ak} dan bapak \textbf{Supri Bin Hj. Amir, S.Si., M.Eng} sebagi pembimbing pertama yang senantiasa  memberikan masukan kepada penulis.
    \item Bapak \textbf{Dr. Hendra, S.Si., M.Kom} dan Ibu \textbf{Nur Hilal, S.Si., M.Si} sebagai tim penguji atas saran dan masukan pada penelitian yang telah dilakukan oleh penulis.
    \item Seluruh Bapak dan Ibu dosen FMIPA Universitas Hasanuddin yang telah mendidik dan memberikan ilmunya sehingga penulis mampu menyelesaikan program sarjana. Serta para staf yang telah membantu dalam pengurusan berkas administrasi.

    \item Saudara-saudara \textbf{Ramsis Squad} \textbf{(Sultan, Muh Rizaldi, Sangereng Dewa Raja, Hajrin, Badaruddin Hidayat, Hamzah Julianto Nugraha, Muh Naim)} sebagai keluarga semasa tinggal di Ramsis sejak menjadi mahasiswa baru, saling berbagi dalam banyak hal, saling membantu dan bahkan saling merepotkan.

>>>>

Ucapan terima kasih dan apresiasi yang tak terhingga kepada kedua orang tua penulis bapak \textbf{Sudarmin} dan ibu \textbf{Yuli Hadiyanti} yang tak kenal lelah dalam memanjatkan doa serta memberikan nasihat dan motivasi kepada penulis. Tak lupa juga kepada saudara-saudara penulis \textbf{Fitri Handayani}, \textbf{Tri Novianti}, \textbf{Jumadil Yusuf}, \textbf{Muhammad Fitrah}, \textbf{Adam Ramadhan} yang selalu menjadi motivasi bagi penulis untuk terus melangkah maju.

@@

\begin{enumerate}[topsep=0pt,itemsep=0pt,partopsep=0pt, parsep=0pt]
    \item Rektor Universitas Hasanuddin, Ibu \textbf{Prof. Dr. Dwia Aries Tina Pulubuhu} beserta jajarannya.
    \item Dekan Fakultas Matematika dan Ilmu Pengetahuan Alam, \textbf{Dr. Eng. Amiruddin} beserta jajarannya.
    \item Ketua Departemen Matematika FMIPA, \textbf{Dr. Nurdin, S.Si., M.Si}, dan juga \textbf{Dr. Muhammad Hasbi, M.Sc} sebagai ketua Program Studi Sistem Informasi Universitas Hasanuddin.
    \item Bapak \textbf{Dr. Eng. Armin Lawi, S.Si., M.Eng} sebagai pembimbing utama yang telah banyak memberikan arahan, ide, motivavsi serta dukungan kepada penulis dalam banyak hal.
    \item Almarhum bapak \textbf{Dr. Diaraya, M.Ak} dan bapak \textbf{Supri Bin Hj. Amir, S.Si., M.Eng} sebagi pembimbing pertama yang senantiasa  memberikan masukan kepada penulis.
    \item Bapak \textbf{Dr. Hendra, S.Si., M.Kom} dan Ibu \textbf{Nur Hilal, S.Si., M.Si} sebagai tim penguji atas saran dan masukan pada penelitian yang telah dilakukan oleh penulis.
    \item Seluruh Bapak dan Ibu dosen FMIPA Universitas Hasanuddin yang telah mendidik dan memberikan ilmunya sehingga penulis mampu menyelesaikan program sarjana. Serta para staf yang telah membantu dalam pengurusan berkas administrasi.

    \item Saudara-saudara \textbf{Ramsis Squad} \textbf{(Sultan, Muh Rizaldi, Sangereng Dewa Raja, Hajrin, Badaruddin Hidayat, Hamzah Julianto Nugraha, Muh Naim, Achmad Husein Nyompa)} sebagai keluarga semasa tinggal di Ramsis sejak menjadi mahasiswa baru, saling berbagi dalam banyak hal, saling membantu dan bahkan saling merepotkan.
